---
title: "Getting Start with De Novo CIDER (dnCIDER): Cross-Species Pancreas Integration"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Start with De Novo CIDER (dnCIDER): Cross-Species Pancreas Integration"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# What is dnCIDER?

**dnCIDER** is a computational method designed for integrative analysis of single-cell RNA-seq data across batches or species. This vignette demonstrates its application to a human-mouse pancreas dataset, showing how to identify conserved cell populations across species.

# Set up

In addition to **CIDER**, we will load the following packages:

```{r setup}
library(CIDER)
library(Seurat)
library(parallel)
library(cowplot)
```

# Load pancreas data

The example data can be downloaded from https://figshare.com/s/d5474749ca8c711cc205. 

Pancreatic cell data$^1$ contain cells from human (8241 cells) and mouse (1886 cells).

```{r} 
# Load counts matrix and metadata
load("../data/pancreas_counts.RData") # count matrix
load("../data/pancreas_meta.RData") # meta data/cell information

# Create Seurat object
seu <- CreateSeuratObject(counts = pancreas_counts, meta.data = pancreas_meta)

# Check batch composition
table(seu$Batch)
```

# Perform dnCIDER (high-level)

DnCIDER contains three steps. 

## Step 1: Initial Clustering

Performs preprocessing and generates initial clusters within each batch:

```{r}
seu <- initialClustering(
  seu,
  additional.vars.to.regress = "Sample",  # Regress out sample-specific effects
  dims = 1:15                             # PCA dimensions to use
)
```

## Step 2: Compute IDER

Estimates batch-corrected similarity matrices:

```{r}
ider <- getIDEr(
  seu,
  downsampling.size = 35,   # Cells per cluster for downsampling
  use.parallel = FALSE,     # Disable parallelization for reproducibility
  verbose = FALSE           # Suppress progress messages
)
```

## Step 3: Final Integrated Clustering

Merges clusters using IDER-derived similarities:

```{r}
seu <- finalClustering(
  seu,
  ider,
  cutree.h = 0.35  # Height for hierarchical clustering cut
)
```

# Visualise clustering results

We use the Seurat pipeline to perform normalisation (`NormalizeData`), preprocessing (`FindVariableFeatures` and `ScaleData`) and dimension reduction (`RunPCA` and `RunTSNE`).

```{r seurat-pipeline}
# Preprocessing for Visualization
seu <- NormalizeData(seu, verbose = FALSE)
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
seu <- ScaleData(seu, verbose = FALSE)
seu <- RunPCA(seu, npcs = 20, verbose = FALSE)
seu <- RunTSNE(seu, reduction = "pca", dims = 1:12)
```

Next we plot integrated clusters vs. ground truth. By comparing the dnCIDER results to the cell annotation from the publication$^1$, we observe that dnCIDER correctly identify the majority of populations across two species.

```{r tsne-plot-CIDER-results, fig.height=3, fig.width=8}
# Generate plots
p1 <- scatterPlot(seu, "tsne", 
                  colour.by = "CIDER_cluster", 
                  title = "Integrated Clusters (dnCIDER)")

p2 <- scatterPlot(seu, "tsne", 
                  colour.by = "Group", 
                  title = "Original Cell Types")

# Arrange side-by-side
plot_grid(p1, p2, ncol = 2)
```

Interpretation: dnCIDER successfully aligns human (prefix h) and mouse (m) cell types. For example:

* Beta cells (hBeta/mBeta) form a unified cluster
* Alpha cells (hAlpha/mAlpha) show cross-species alignment
* Minor populations like Acinar and Ductal are conserved

# Notes & Best Practices

1. **Downsampling Size**: Adjust `downsampling.size` (default: 35) if clusters are small.
1. **Batch Variable**: Ensure your Seurat object contains a batch identifier (default column name: `"Batch"`).
1. **Visualization**: Always validate integration using known marker genes in addition to embeddings.
1. **Runtime**: For large datasets, enable parallelization with `use.parallel = TRUE`.

# Reproducibility

```{r sessionInfo}
sessionInfo()
```

# References

1. Baron, M. et al. A Single-Cell Transcriptomic Map of the Human and Mouse Pancreas Reveals Inter- and Intra-cell Population Structure. Cell Syst 3, 346–360.e4 (2016).
2. Satija R, et al. Spatial reconstruction of single-cell gene expression data. Nature Biotechnology 33, 495-502 (2015). 

